On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-29 paper/summary.tex
rough version 3  by MKL

\section{Summary and discussion}\label{summary}
In this paper, we develop the linear theory of cooling, irradiated,
and viscous accretion disks in order to understand gravitational
instability (GI) in realistic protoplanetary disks (PPDs). 
We use a Navier-Stokes viscosity to mimic the effects of
turbulent angular momentum transport.
%KMK
. %which may be of any origin, though we specialize to
            %gravito-turb 
This viscosity provides a background heating to balance the imposed cooling,  
and may also act on linear perturbations. %thermo and dynamic 
We suggest that disk fragmentation, observed in numerical simulations, 
can be understood as the eventual outcome of secondary
instabilities of a gravito-turbulent base state, driven by cooling
and/or viscosity.

%KMK
Previous work has focused on the impact of viscosity and 
cooling on the equilibrium temperature of the disk, but merely used this 
to measure the classic Toomre Q and thereby assess stability. We demonstrate
through a more rigorous inclusion of these effects into the dispersion relation that
they can drive secondary instabilities.
While viscosity and cooling can be related through thermal
balance, they independently enhance growth rates: cooling reduces
thermal stabilization; and viscous forces compromise rotational
stabilization. This provides a physical explanation as to why
rapidly-cooled, gravito-turbulent disks cannot exist. Moreover, we discuss below how
these models may lend support to the varied behavior observed in numerical simulations.
%KMK

@@

The effect of cooling and irradiation on GI is quantified by the 
dispersion relation 
Eq. \ref{inviscid}. We find a sufficient condition for stability which  
depends on the irradiation level (Eq. \ref{stable_condition}) but is
\emph{independent} of the cooling time. 
This means that long cooling times can still  
formally lead to instability. However, growth timescales may be 
uninterestingly long for cooling times $\tcool\Omega\gtrsim 
O(10)$. Because cooling affects pressure support, GI driven by cooling
occur on small scales, $kH\gtrsim O(1)$.

@@

We apply our linear framework to protoplanetary disks with 
realistic models for cooling and gravito-turbulence. 
We show that with a physically motivated cooling model for PPDs, 
cooling alone does not lead to gravitational instabilities. This is due
to stabilization by an effective `irradiation' associated with the PPD
cooling function, which is present even if 
there is no physical irradiation. Physically, this  term
represents the fact that disks do not cool towards zero temperature because
radiative cooling is a strong (quartic) function of temperature  %%KMK?

%due to matching functional forms of
                                %cooling 
Instead, viscous GI occurs on  
dynamical timescales for $R\gtrsim 60$AU because 
$\alpha\gtrsim0.1$ there. This corresponds to a Toomre $Q\simeq 1.5$ and 
a cooling time $\tcool\lesssim 3\Omega^{-1}$. These are coincident with 
\emph{empirical} conditions cited in the literature to determine disk
fragmentation 
\citep[e.g.][]{rafikov15}. Here, we attribute a \emph{physical}
cause for the fragmentation of realistic PPDs: gravito-turbulent PPDs
fragment when turbulent stresses are large enough to further
destabilize the disk against self-gravity.    
%collary: details of cooling unimportant as long as the associated
%stress is the same

@@

\subsection{Relation to numerical simulations}\label{prev_works}
%frag in sims because of : cooling, viscosity, or high den
%flucuations? don't know 
Our results may shed light on some numerical simulations 
concerning disk fragmentation. Table 
\ref{bstar_compare} shows a close match between the characteristic
cooling time for cooling-driven GI (Eq. \ref{betastar}) 
and that for disk fragmentation observed in simulations 
\citep{gammie01,rice05,rice11}. This suggests that, at least for those
simulations, fragmentation is physically due to the removal of
pressure stabilization by cooling on a radial scale of the disk scale
height.
%%KMK
%Note that unlike our physical cooling function (Eq. \ref{real_cool}), the standard beta function in these
%numerical simulations cools towards zero temperature, rather than a temperature floor or irradiation temperature 
%%KMK some comment like that here, or too redundant with comment below??
 %i.e. cool-driven g 
{\bf don't know if we want to dwell on this table}
%negligible effect from viscous heating
%makes sense? sims show clumps are not turbulent internally
%then visc plays no role in clumping -> only a cooling effect

@@

We suggest having a floor temperature in the standard beta
cooling prescription is a necessary condition for  
numerical convergence. This limits the
relevant scales to a finite range.  %for cool gi 
%physically irradiation 
Furthermore, without a floor temperature, standard beta cooling is
a function of the pressure/energy density field only. It is not clear
how relevant are results obtained from this is to actual PPDs,
where cooling depends on two thermodynamic states (e.g. pressure and
temperature). A floor temperature permits a mapping between standard beta
cooling and PPD cooling (\S\ref{ppd_cooling}). 
%KMK
Moreover, the standard beta cooling does 
not account for optical depth effects, which force the midplane 
and high density perturbations to cool more slowly.

@@

%KMK
A temperature floor may be a necessary, but not sufficient condition
for numerical convergence. \cite{baehr15} included a floor temperature 
in their simulations but still find that at fixed cooling rates, disks
eventually fragment with sufficient spatial resolution.
In light of our results on viscous GI, 
we suggest another possible contribution to non-convergence: %in addition to num visc (meru lodato)
high resolution enables small-scale turbulent angular momentum transport 
% physical, cf numerical  
to aid clump formation via the removal of rotational support. (see also \S\ref{MHD} below). 
If clumps are only marginally resolved, simulations may not have sufficient dynamic range for turbulent eddies
to cascade down to these scales.
%This mechanism is unavailable at low resolutions, for which one must
%rely on rapid cooling.  
%maybe there is an effective floor temp which depends on res

We comment that although modern simulations resolve the dominant
scale associated with gravito-turbulence very well, $kH\sim 1$
\citep{cossins09}, this does not
necessarily imply small-scale dynamics/thermodynamics are unimportant 
\citep[especially for non-linear evolution, ][]{young15},  
as is evident from the non-converging simulations described above.   
%KMK
Even at very high resolution one might worry that the artificial dissipation scale
imposed by the grid / smoothing length is too similar to scale of fragmentation.
%it is true that most unstable wavelength is H is this is always
%resolved
%but that doesn't mean small-scale structures are unimportant 
%note: no turbulence observed inside clumps
%consistent with rho*nu = const  
%this alpha should be that associated with small-scale transport 
%frag due to cooling -> structures less than H
%frag due to viscosity -> structures more than H
%stochastic frag 
%non-convergence with resolution 
%no magic beta when non-ideal effects are included

\subsection{MHD turbulence}\label{MHD}%turbulence and viscosity 
We emphasize the linear framework we have developed does not assume  
a particular origin for the turbulence that is represented by the imposed
viscosity. For example, the 2D dispersion relation given by Eq. \ref{thindisk}
with Eq. \ref{bigA}---\ref{bigF}, treats $\alpha$ as an independent
input parameter.

@@

Our deterministic approach cannot model `stochastic fragmentation'
\citep{paardekooper12,hopkins13}. In this interpretation,  
fragmentation is attributed to the occurence and survival 
of large, non-linear density enhancements, which arise from the
gravito-turbulent flucuations simply by chance. %KMK
There is insufficient evidence that gravito-turbulence adequately samples
the density power spectrum as assumed by \cite{hopkins13}.
%cooling affects survival rate 
Nevertheless, one might consider this form of fragmentation as a secondary instability
triggered by lowering the local
Toomre parameter through a (random) increase in density 
(Eq. \ref{toomreQ_criterion}). By contrast, the viscosity/cooling
destabilizing mechanisms discussed in this paper lowers $Q$ by
reducing pressure and Coriolis forces. 
{\bf KMK: this last sentence sounds like it's just about changing classic Q}.
%we have a physical way of lowering pressure/rotation

@@

%modeling turb as visc %explore other visc laws 
The most important assumption in this work is modeling turbulence as a  
Navier-Stokes viscosity. Furthermore, we have chosen a particular  
viscosity law (see \S\ref{visc_model}) to mimic the effect of 
turbulence in reducing rotational support (in the sense that it
provides small-scale angular momentum transport). 
%to enhance clumping 
%this visc law is used in prev studies (by much smarter people)
How to quantitatively model gravito- or MHD turbulence 
as a viscosity, especially on dynamical timescales, should be 
clarified with direct numerical simulations. The present viscosity
models should then be modified accordingly. 
%could consider other visc parameters mu and lambda (but this
%introduces visc overstability, or new functional forms of the
%viscosity law altogether

>>>>

\section{Summary and discussion}\label{summary}
In this paper, we develop the linear theory of cooling, irradiated, 
and viscous accretion disks in order to understand gravitational 
instability (GI) in realistic protoplanetary disks (PPDs). 
We use a Navier-Stokes viscosity to mimic the effects of
turbulent angular momentum transport. 
%KMK
%which may be of any origin, though we specialize to
            %gravito-turb 
This viscosity provides a background heating to balance the imposed cooling,  
and may also act on linear perturbations. %thermo and dynamic 
We suggest that disk fragmentation, observed in numerical simulations, 
can be understood as the eventual outcome of secondary
instabilities of a gravito-turbulent base state, driven by cooling
and/or viscosity. 
%frag in sim and frag in real disk may or may not have the same
%physical origin

%KMK
Previous work has focused on the impact of viscosity and 
cooling on the equilibrium temperature and surface density of an 
accretion disk, but merely used this to  the classic Toomre $Q$,
thereby assess stability. %Disk fragmentation is assessed similarly,
%but is usually based on empirically-derived conditions (see below). 
We demonstrate by explicitly including these effects into the
dispersion relation that they can drive secondary instabilities.
While viscosity and cooling can be related through thermal
balance, they independently enhance growth rates: cooling reduces
thermal stabilization; and viscous forces compromise rotational
stabilization. This provides a physical explanation as to why
rapidly-cooled, gravito-turbulent disks cannot exist. Moreover, we discuss below how
these models may lend support to the varied behavior observed in numerical simulations.
%KMK

@@

The effect of cooling and irradiation on GI is quantified by the 
dispersion relation Eq. \ref{inviscid}. 
We find sufficient conditions for stability which  
depends on the irradiation level (Eq. \ref{stable_condition}) but is 
\emph{independent} of the cooling time. 
This means that long cooling times can still  
formally lead to instability. However, growth timescales may be 
uninterestingly long for cooling times $\tcool\Omega\gtrsim 
O(10)$. Because cooling affects pressure support, GI driven by cooling
occur on small scales, $kH\gtrsim O(1)$.

@@

We apply our linear framework to protoplanetary disks with 
realistic models for cooling and gravito-turbulence. 
We show that with a physically motivated cooling model for PPDs, 
cooling alone does not lead to gravitational instabilities. This is
due to stabilization by an effective `irradiation' associated with the
PPD cooling function, which is present even if there is no physical
irradiation. 
This effective irradiation arises because, for the stability problem,
%cf eqm disk 
a non-zero floor temperature is generally required in the cooling
prescription (Eq. \ref{beta_cool}) in order to map it to PPD cooling,
regardless of the actual irradiation temperature.

%it's associated with density-dependence

%is because PPD cooling depends on both pressure and
%temperature. Mapping this 
%to the standard beta cooling function requires a non-zero floor
%temperature, otherwise standard beta c

%This can only be mapped to the
%standard beta cooling 
%Physically, this  term
%represents the fact that disks do not cool towards zero temperature because
%radiative cooling is a strong (quartic) function of temperature  %%KMK?
%due to matching functional forms of
                                %cooling

Instead, viscous GI occur on dynamical timescales in a PPD for
$R\gtrsim 60$AU because $\alpha\gtrsim0.1$ there. 
This corresponds to a Toomre $Q\simeq 1.5$ and 
a cooling time $\tcool\lesssim 3\Omega^{-1}$. These are coincident with 
\emph{empirical} conditions cited in the literature to determine disk
fragmentation 
\citep[e.g.][]{rafikov15}. Here, we attribute a \emph{physical}
cause for the fragmentation of realistic PPDs: gravito-turbulent PPDs
fragment when turbulent stresses are large enough to further
destabilize the disk against self-gravity.    
%collary: details of cooling unimportant as long as the associated
%stress is the same

@@

\subsection{Relation to numerical simulations}\label{prev_works}
%frag in sims because of : cooling, viscosity, or high den
%flucuations? don't know 
Our results may help understand some numerical simulations 
concerning disk fragmentation. Table 
\ref{bstar_compare} shows a close match between the characteristic
cooling time for cooling-driven GI (Eq. \ref{betastar}) 
and that for disk fragmentation observed in simulations 
\citep{gammie01,rice05,rice11}. This suggests that, at least for those
simulations, fragmentation is physically due to the removal of
thermal stabilization by cooling on radial lengthscales of the disk 
thickness.   
%i.e. there is no effective heating mechanism on the scale of
        %H 
%{\bf This would imply over-densities on this scale 
%do not experience heating during its growth. This is not unreasonable
%because in reality, gravito-turbulent heating only occurs when a 
%spiral shock passes through said over-density \citep{young16}. Thus
%there there is a time interval during which only cooling acts, whence
%our inviscid theory can be applied. --- too much?}

%%KMK
%Note that unlike our physical cooling function (Eq. \ref{real_cool}),
%the standard beta function in these 
%numerical simulations cools towards zero temperature, rather than a
%temperature floor or irradiation temperature  
%%KMK some comment like that here, or too redundant with comment below??
 %i.e. cool-driven g

%{\bf don't know if we want to dwell on this table}
%negligible effect from viscous heating
%makes sense? sims show clumps are not turbulent internally
%then visc plays no role in clumping -> only a cooling effect

@@

We suggest having a floor temperature in the standard beta
cooling prescription is a necessary for  
numerical convergence. This limits the
relevant scales to a finite range.  %for cool gi 
%physically irradiation 
Furthermore, without a floor temperature, standard beta cooling is
a function of the pressure/energy density field only. It is not clear
how relevant are results obtained from this is to actual PPDs,
where cooling depends on two thermodynamic states (e.g. pressure and
temperature). A floor temperature permits a mapping between standard beta
cooling and PPD cooling (\S\ref{ppd_cooling}). 
%KMK
Moreover, the standard beta cooling does 
not account for optical depth effects, which force the midplane 
and high density perturbations to cool more slowly.

@@

%KMK
A temperature floor may be a necessary, but not sufficient condition
for numerical convergence. \cite{baehr15} included a floor temperature 
in their simulations but still find that at fixed cooling rates, disks
eventually fragment with sufficient spatial resolution.
In light of our results on viscous GI, 
we suggest another possible contribution to non-convergence: %in addition to num visc (meru lodato)
high resolution enables small-scale turbulent angular momentum transport 
% physical, cf numerical  
to aid clump formation via the removal of rotational support. (See also \S\ref{MHD} below.) 
If clumps are only marginally resolved, simulations may not have sufficient dynamic range for turbulent eddies
to cascade down to these scales.
%This mechanism is unavailable at low resolutions, for which one must
%rely on rapid cooling.  
%maybe there is an effective floor temp which depends on res

We comment that although modern simulations resolve the dominant
scale associated with gravito-turbulence very well, $kH\sim 1$
\citep{cossins09}, this does not
necessarily imply small-scale dynamics/thermodynamics are unimportant 
\citep[especially for non-linear evolution, ][]{young15},  
as is evident from the non-converging simulations described above.   
%KMK
Even at very high resolution one might worry that the artificial dissipation scale
imposed by the grid/smoothing length is too similar to scale of fragmentation.
%it is true that most unstable wavelength is H is this is always
%resolved
%but that doesn't mean small-scale structures are unimportant 
%note: no turbulence observed inside clumps
%consistent with rho*nu = const  
%this alpha should be that associated with small-scale transport 
%frag due to cooling -> structures less than H
%frag due to viscosity -> structures more than H
%stochastic frag 
%non-convergence with resolution 
%no magic beta when non-ideal effects are included

\subsection{MHD turbulence}\label{MHD}%turbulence and viscosity 
We emphasize the linear framework we have developed does not assume  
a particular origin for the turbulence that is represented by the imposed
viscosity. For example, our 2D dispersion 
relation, Eq. \ref{thindisk} with Eq. \ref{bigA}---\ref{bigF}, treats 
$\alpha$ as an independent input parameter.

@@

Our deterministic approach cannot model `stochastic fragmentation'
\citep{paardekooper12,hopkins13}. In this interpretation,  
fragmentation is attributed to the occurence and survival 
of large, non-linear density enhancements, which arise from the
gravito-turbulent flucuations simply by chance. %KMK
There is insufficient evidence that gravito-turbulence adequately samples
the density power spectrum as assumed by \cite{hopkins13}.
%cooling affects survival rate 
Nevertheless, one might consider this form of fragmentation as a secondary instability
triggered by lowering the local 
Toomre parameter through a (random) increase in density 
(Eq. \ref{toomreQ_criterion}).
% By contrast, the viscosity/cooling
%destabilizing mechanisms discussed in this paper lowers $Q$ by
%reducing pressure and Coriolis forces. 
%{\bf KMK: this last sentence sounds like it's just about changing classic Q}.
%we have a physical way of lowering pressure/rotation

@@

%modeling turb as visc %explore other visc laws 
The most important assumption in this work is modeling turbulence as a  
Navier-Stokes viscosity. Furthermore, we have chosen a particular  
viscosity law (see \S\ref{visc_model}) to mimic the effect of 
turbulence in reducing rotational support (in the sense that it
provides small-scale angular momentum transport). 
%to enhance clumping 
%this visc law is used in prev studies (by much smarter people)
How to quantitatively model the effect of gravito- or MHD turbulence  
as a viscosity, especially on dynamical timescales, should be 
clarified with direct numerical simulations. The present viscosity
models should then be modified accordingly. 
%could consider other visc parameters mu and lambda (but this
%introduces visc overstability, or new functional forms of the
%viscosity law altogether
